### 2016-07-04 ansi-isolation.tex
various edits to regain space
typo fixes mostly

Fig.~\ref{fig:rel-defs} shows various relations defined over elements
in a trace. In the context of a trace $(\A,\visZ)$, an effect $\eta$
is said to belong to a transaction $T_i$ if $\eta$ belongs to the
effect set $A$ and its transaction identifier is $T_i$. The
containment relation is trivially lifted to the set of effects to
define $\underE{S \subseteq T_i}$.  Visibility and session order
relations are denoted by $\eta_1 \visar \eta_2$ and $\eta_1 \soar
\eta_2$, respectively. A transaction $T_i$ is said to be visible to an
effect $\eta$ if every effect $\eta_1$ of $T_i$ recorded by the trace
is visible to $\eta$.  $T_i$ may be visible to $\eta$ but may not be
visible to every other effect in the $\txn(\eta)$. For a transaction
$T_i$ to be considered to be visible to a transaction $T_j$ in the
context of a trace $\E$ (written $\underE{T_i \visar T_j}$), every
effect ($\eta_1$) of $T_i$ present in $\E$ must be visible to every
effect ($\eta_2$) of $T_j$ in $\E$.  Conversely, if none of the
effects of $T_i$ present in $\E$ are visible to any effect of $T_j$,
then $T_i$ is considered invisible to $T_j$ under $\E$ (written
$\underE{T_i \invisar T_j}$). Transaction $T_i$ is said to have
written to a variable $X$ under $\E$ (i.e., $\underE{T_i \wrstoar X}$)
if there exists a write-to-$X$ effect from $T_i$ in $\E$.
\emph{Reads-from} ($\underE{T_i \rdsfmar X}$) is defined similarly.
$T_i \usesar X$ if it reads or writes $X$.

Fig.~\ref{fig:ansi-isolation} shows various isolation guarantees
defined as propositions indexed by transaction identifiers.
Transaction $T_j$ is said to have experienced \emph{read-my-writes}
visibility under $\E$ if every effect ($\eta_1$) of $T_j$ is visible
to every subsequent effect ($\eta_2$) of the same transaction in $\E$.
This lets $T_j$ to never lose its own updates. Monotonic visibility
adds one more constraint to read-my-writes; besides requiring $\eta_1$
to be visible to $\eta_2$, it also requires every effect ($\eta$)
visible to $\eta_1$ in $\E$ to be visible to $\eta_2$ as well. Thus,
later operations in a transaction witness at least the same set of
effects witnessed by the earlier operations, if not more (hence,
``monotonic'' visibility). Atomic visibility allows an effect
$\eta_2$ of $T_j$ to witness an effect $\eta_1$ of $T_i$ only if all
effects of $T_i$ in $\E$ are also visible to $\eta_2$. Atomic
visibility thus prevents a transaction from being partially visible.
However, atomic visibility does not prevent an uncommitted transaction
from being visible. This is addressed by \C{CommitVis}, which requires
the commit effect of $T_i$ to be visible whenever any effect of $T_i$
is visible.

The ANSI SQL 92 standard requires \iso{Read Committed} isolation to
avoid the dirty reads phenomenon, which is achieved by enforcing
$\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ guarantees. The {\sc rc}
specification\footnote{Note the distinction in fonts: {\sc rc} is an
abbreviation of \iso{Read Committted}, whereas \C{RC} is its
specification.} (\C{RC}) is therefore a combination of these two
guarantees. The specification also agrees with the description and
implementation~\cite{bailishat,pldi15} of {\sc rc} for highly
available replicated stores. On relational databases, however, {\sc
rc} has also come to be associated with the $\mathtt{MonotonicVis}$
guarantee.  Nonetheless, $\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$
are sufficient to reason about {\sc rc} isolation on relational stores
too. The combination of these guarantees with the {\sc sc} property of
relational stores (formalized in \S\ref{sec:store-consistency})
automatically leads to the monotonicity guarantee, which explains why
{\sc rc} comes with $\mathtt{MonotonicVis}$ on such stores regardless
of the implementation. On weakly consistent stores however,
$\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ do not imply
$\mathtt{MonotonicVis}$. A stronger isolation level called
\iso{Monotonic Atomic View} ({\sc mav} of
Fig.~\ref{fig:ansi-isolation})~\cite{bailishat} was proposed to
explicitly extend {\sc rc} with monotonicity on such stores.

@@

\caption{Sample executions of concurrent transactions at different
isolation levels. Solid (black) arrows indicate the timeline. Points
on the timeline mark the time when an operation is executed. Dotted
(blue) arrows denote $\visZ$. }
\label{fig:ansi-iso-eg}
\end{figure}

Snapshot visibility (\C{SnapshotVis}) captures the scenario where a
transaction executes against a static snapshot of the database.  A
transaction $T_i$ is said to be snapshot-visible to a transaction
$T_j$ if either it is visible to $T_j$ (i.e., $T_i$ is included in the
snapshot), or it is invisible (i.e., it is not included); it is
forbidden for only a suffix (more generally, a subset) of $T_j$ to
witness $T_i$. Specification of ANSI SQL \iso{Repeatable Read}
isolation (\C{RR}) extends \C{MAV} with the snapshot visibility
guarantee.  Observe that snapshot visibility permits $T_i$ and $T_j$
to execute and commit while being oblivious of each other. This
scenario is captured in Fig.~\ref{fig:ansi-iso-eg-rr}, where
transactions $T_1$ and $T_2$, which perform conflicting writes,
execute against a snapshot of the database and commit concurrently.
While the actual values read and written by $T_1$ and $T_2$ are
unimportant (hence, elided), it is important to note the
\emph{absence} of visiblity arrows from $T_2$ to $T_1$, although $T_2$
commits before $T_1$'s read-from-$Y$.

\iso{Snapshot Isolation} ({\sc si}) proscribes this possibility. If
$T_i$ and $T_j$ both write to the same shared variable, then {\sc si}
insists that either $T_i$ be visible to $T_j$, or $T_j$ be visible to
the conflicting write of $T_i$ (this is captured by the auxiliary
definition \C{SnapshotSER}). Fig.~\ref{fig:ansi-iso-eg-si} shows an
execution where {\sc si} transaction $T_1$ witnesses a snapshot of the
database that doesn't include $T_2$. Due to \C{SI($T_1$)}, the
conflicting write to $X$ in $T_2$ is now required to witness $T_1$, as
captured by the direction of the $\visZ$ arrow (subsequent operations
also witness $T_1$ because $T_2$ is executing under {\sc mav}).
Lastly, \iso{Serializable} isolation extends $\mathtt{SnapshotSER}$ to
also cover non-conflicting transactions that write to variables read
in the current transaction.  Fig.~\ref{fig:ansi-iso-eg-ser} shows an
execution of an {\sc ser} transaction $T_3$ and a {\sc mav}
transaction $T_4$, which write to $X$ and $Y$, respectively. The
snapshot witnessed by the {\sc ser} transaction $T_3$ does not include
$T_4$, but the write to $Y$ in $T_4$, although non-conflicting,
witnesses $T_1$ because $Y$ is read by $T_1$. Visibility includes
other operations of $T_4$ because of {\sc mav}. Note that {\sc ser}
does not guarantee total order of transactions w.r.t. $\visZ$ because,
in reality, databases do not guarantee serializable execution unless
\emph{all} transactions choose {\sc ser}.

>>>>

Fig.~\ref{fig:rel-defs} shows various relations defined over elements
in a trace. In the context of a trace $(\A,\visZ)$, an effect $\eta$
is said to belong to a transaction $T_i$ if $\eta$ belongs to the
effect set $A$ and its transaction identifier is $T_i$. The
containment relation is trivially lifted to the set of effects to
define $\underE{S \subseteq T_i}$.  Visibility and session order
relations are denoted by $\eta_1 \visar \eta_2$ and $\eta_1 \soar
\eta_2$, respectively. A transaction $T_i$ is said to be visible to an
effect $\eta$ if every effect $\eta_1$ of $T_i$ recorded by the trace
is visible to $\eta$.  $T_i$ may be visible to $\eta$ but may not be
visible to every other effect in the $\txn(\eta)$. For a transaction
$T_i$ to be considered to be visible to a transaction $T_j$ in the
context of a trace $\E$ (written $\underE{T_i \visar T_j}$), every
effect ($\eta_1$) of $T_i$ present in $\E$ must be visible to every
effect ($\eta_2$) of $T_j$ in $\E$.  Conversely, if none of the
effects of $T_i$ present in $\E$ are visible to any effect of $T_j$,
then $T_i$ is considered invisible to $T_j$ under $\E$ (written
$\underE{T_i \invisar T_j}$). Transaction $T_i$ is said to have
written to a variable $X$ under $\E$ (i.e., $\underE{T_i \wrstoar X}$)
if there exists a writes-to-$X$ effect from $T_i$ in $\E$.
\emph{Reads-from} ($\underE{T_i \rdsfmar X}$) is defined similarly.
$T_i \usesar X$ holds if $T_i$ reads or writes $X$.

Fig.~\ref{fig:ansi-isolation} shows various isolation guarantees
defined as propositions indexed by transaction identifiers.
Transaction $T_j$ is said to have experienced \emph{read-my-writes}
visibility under $\E$ if every effect ($\eta_1$) of $T_j$ is visible
to every subsequent effect ($\eta_2$) of the same transaction in $\E$.
This lets $T_j$ to never lose its own updates. Monotonic visibility
adds one more constraint to \emph{read-my-writes}; besides requiring $\eta_1$
to be visible to $\eta_2$, it also requires every effect ($\eta$)
visible to $\eta_1$ in $\E$ to be visible to $\eta_2$ as well. Thus,
later operations in a transaction witness at least the same set of
effects witnessed by the earlier operations, if not more (hence,
``monotonic'' visibility). Atomic visibility allows an effect
$\eta_2$ of $T_j$ to witness an effect $\eta_1$ of $T_i$ only if all
effects of $T_i$ in $\E$ are also visible to $\eta_2$. Atomic
visibility thus prevents a transaction from being partially visible.
However, atomic visibility does not prevent an uncommitted transaction
from being visible. This is addressed by \C{CommitVis}, which requires
the commit effect of $T_i$ to be visible whenever any effect of $T_i$
is visible.

The ANSI SQL 92 standard requires \iso{Read Committed} isolation to
avoid the dirty reads phenomenon, which is achieved by enforcing
$\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ guarantees. The {\sc rc}
specification (\C{RC})\footnote{In the following, we use small caps to abbreviate
  isolation levels (\emph{e.g.,} {\sc rc} for \emph{Read Committed}), and
  typewriter font for abbreviations of the specification of an isolation
  level (\emph{e.g.,} \C{RC} denotes the specification for {\sc rc} as
  given in Fig.~\ref{fig:ansi-isolation}).} is therefore a combination of these two
guarantees. The specification also agrees with the description and
implementation~\cite{bailishat,pldi15} of {\sc rc} for highly
available replicated stores. On relational databases, however, {\sc
rc} has also come to be associated with the $\mathtt{MonotonicVis}$
guarantee.  Nonetheless, $\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$
are sufficient to reason about {\sc rc} isolation on relational stores
too. The combination of these guarantees with the {\sc sc} property of
relational stores (formalized in \S\ref{sec:store-consistency})
automatically leads to the monotonicity guarantee, which explains why
{\sc rc} comes with $\mathtt{MonotonicVis}$ on such stores regardless
of the implementation. On weakly consistent stores however,
$\mathtt{AtomicVis}$ and $\mathtt{CommitVis}$ do not imply
$\mathtt{MonotonicVis}$. A stronger isolation level called
\iso{Monotonic Atomic View} ({\sc mav} of
Fig.~\ref{fig:ansi-isolation})~\cite{bailishat} was proposed to
explicitly extend {\sc rc} with monotonicity on such stores.

Snapshot visibility (\C{SnapshotVis}) captures the scenario where a
transaction executes against a static snapshot of the database.  A
transaction $T_i$ is said to be snapshot-visible to a transaction
$T_j$ if either it is visible to $T_j$ (i.e., $T_i$ is included in the
snapshot), or it is invisible (i.e., it is not included); it is
forbidden for only a suffix (more generally, a subset) of $T_j$ to
witness $T_i$. The specification of ANSI SQL \iso{Repeatable Read}
isolation (\C{RR}) extends \C{MAV} with the snapshot visibility
guarantee.  Observe that snapshot visibility permits $T_i$ and $T_j$
to execute and commit while being oblivious of each other. This
scenario is captured in Fig.~\ref{fig:ansi-iso-eg-rr}, where
transactions $T_1$ and $T_2$, which perform conflicting writes,
execute against a snapshot of the database and commit concurrently.
While the actual values read and written by $T_1$ and $T_2$ are
unimportant (hence, elided), it is important to note the
\emph{absence} of visiblity arrows from $T_2$ to $T_1$, although $T_2$
commits before $T_1$'s read-from-$Y$.

\iso{Snapshot Isolation} ({\sc si}) proscribes this possibility. If
$T_i$ and $T_j$ both write to the same shared variable, then {\sc si}
insists that either $T_i$ be visible to $T_j$, or $T_j$ be visible to
the conflicting write of $T_i$ (this is captured by the auxiliary
definition \C{SnapshotSER}). Fig.~\ref{fig:ansi-iso-eg-si} shows an
execution where {\sc si} transaction $T_1$ witnesses a snapshot of the
database that doesn't include $T_2$. Due to \C{SI($T_1$)}, the
conflicting write to $X$ in $T_2$ is now required to witness $T_1$, as
captured by the direction of the $\visZ$ arrow (subsequent operations
also witness $T_1$ because $T_2$ is executing under {\sc mav}).
Lastly, \iso{Serializable} isolation extends $\mathtt{SnapshotSER}$ to
also cover non-conflicting transactions that write to variables read
in the current transaction.  Fig.~\ref{fig:ansi-iso-eg-ser} shows an
execution of an {\sc ser} transaction $T_3$ and a {\sc mav}
transaction $T_4$, which write to $X$ and $Y$, respectively. The
snapshot witnessed by the {\sc ser} transaction $T_3$ does not include
$T_4$, but the write to $Y$ in $T_4$, although non-conflicting,
witnesses $T_1$ because $Y$ is read by $T_1$. Visibility includes
other operations of $T_4$ because of {\sc mav}. Note that {\sc ser}
does not guarantee a total order on transactions w.r.t. $\visZ$
because, in practice, databases do not guarantee global serializable
execution unless \emph{all} transactions choose {\sc ser}.

@@

\caption{\small Sample executions of concurrent transactions at different
isolation levels. Solid (black) arrows indicate the timeline. Points
on the timeline mark the time when an operation is executed. Dotted
(blue) arrows denote $\visZ$. }
\label{fig:ansi-iso-eg}
\vspace*{-12pt}
\end{figure}

### 2016-07-04 case-studies.tex
typo fixes mostly

\paragraph{A video view counter.} Our first application is a
Ruby-on-Rails implementation of a counter to count the number of views
a video has garnered on a video-sharing website like
YouTube\footnote{Source code of the applications, experiments, and the
  tools we developed to analyze Ruby can be found at the following
  anonymized url: \url{...}}. The application stores the video count
in a MySQL database, and supports a read operation (a database read
transaction), and an increment operation (a read-increment-write
sequence wrapped in a transaction). The application is expected to
preserve the \emph{monotonicity invariant}, namely that the view count
should never appear to be decreasing. However, the default
\iso{Repeatable Read} isolation level of MySQL leads to anomalous
executions - with 32 concurrent writers performing single increments,
and one reader issuing a constant stream of reads, we observed 12
violations of monotonicity in 10 rounds. We subsequently instantiated
our reasoning framework for an {\sc sc} store with ANSI SQL isolation
levels, and spent 5 man hours determining appropriate isolation levels
needed to enforce the invariant\footnote{The representative \txnimp
  programs we used to formally reason about scenarios described in
  this section can be found in the supplementary.} In particular, we
were able to formally prove that to preserve the invariant, writers
need to execute at \iso{Serializable} isolation, while readers can
execute at \iso{Read Uncommitted}. Repeating the experiments with this
configuration led to no violations, as expected. While executing all
transactions at {\sc ser} also prevents violations, it led to an
additional 113\% increase in the latency of writes on an average (32
concurrent writers; 5 rounds). We repeated the experiments with the
same configuration on Postgres, and found no violations. However,
Postgres's isolation levels~\cite{postgresiso} are stronger than ANSI
SQL's in non-trivial ways. We therefore formalized Postgres'
guarantees and repeated relevant parts of the proof to discover that
\iso{Repeatable Read} is in fact sufficient for writers - an
observation we confirmed through experiments. As usual, using {\sc
  ser} for all transactions also suffices, but on Postgres results in
an additional 47\% increase in write latency.

@@

To understand the effect of changing the store consistency levels, we

@@

monotonically increasing count\footnote{The trade-off is that the view
count is now no longer accurate, although it grows monotonically.}.

@@

site, also written using Ruby-on-Rails~\cite{railsbook}. One of the
invariants that the application chooses to enforce is referential

@@

microposts authored by the user.  However, concurrent observers can
nonetheless witness violations of referential integrity. To measure

@@

which implements ANSI SQL isolation levels. Repeating our experiments

@@

expectected), while yielding a 38\% reduction in latency than an {\sc

@@

Postgres's isolation levels are stronger than the corresponding ANSI

>>>>

\paragraph{A video view counter.} Our first application is a
Ruby-on-Rails implementation of a counter to count the number of views
a video has garnered on a video-sharing website like
YouTube\footnote{Source code of the applications, experiments, and the
  tools we developed to analyze Ruby can be found at the following
  anonymized url: \url{...}}. The application stores the video count
in a MySQL database, and supports a read operation (a database read
transaction), and an increment operation (a read-increment-write
sequence wrapped in a transaction). The application is expected to
preserve the \emph{monotonicity invariant}, namely that the view count
should never appear to be decreasing. However, the default
\iso{Repeatable Read} isolation level of MySQL leads to anomalous
executions - with 32 concurrent writers performing single increments,
and one reader issuing a constant stream of reads, we observed 12
violations of monotonicity in 10 rounds. We subsequently instantiated
our reasoning framework for an {\sc sc} store with ANSI SQL isolation
levels, and spent 5 man hours determining appropriate isolation levels
needed to enforce the invariant.\footnote{The representative \txnimp
  programs we used to formally reason about scenarios described in
  this section can be found in the supplementary.} In particular, we
were able to formally prove that to preserve the invariant, writers
need to execute using \iso{Serializable} isolation, while readers can
execute using \iso{Read Uncommitted}. Repeating the experiments with
this configuration led to no violations, as expected. While executing
all transactions at {\sc ser} also prevents violations, it led to an
additional 113\% increase in the latency of writes on average, using a
configuration with 32 concurrent writers, and 5 rounds. We repeated
the experiments with the same configuration on Postgres, and found no
violations. However, Postgres' isolation levels~\cite{postgresiso} are
stronger than ANSI SQL's in non-trivial ways. We therefore formalized
Postgres' guarantees and repeated relevant parts of the proof to
discover that \iso{Repeatable Read} is in fact sufficient for writers
- an observation confirmed through experiments. As usual, using {\sc
  ser} for all transactions also suffices, but on Postgres results in
an additional 47\% increase in write latency.

@@

To understand the effect of changing store consistency levels, we

@@

monotonically increasing count.\footnote{The trade-off is that the view
count is now no longer accurate, although it grows monotonically.}

@@

site, also written using Ruby-on-Rails~\cite{railsbook}.  An important
invariant that must be preserved is referential

@@

microposts authored by the user.  However, it is possible for concurrent observers 
to nonetheless witness invariant violations. To measure

@@

which implements ANSI SQL isolation. Repeating our experiments

@@

expected), while yielding a 38\% reduction in latency than an {\sc

@@

Postgres isolation levels are stronger than the corresponding ANSI

### 2016-07-04 introduction.tex
typo fixes mostly

The paper makes the following contributions:
\begin{itemize}
  \item We develop a semantics for a core language equipped with
    weakly-isolated transactions, demonstrating that a general
    parametric isolation semantics can be expressed as
    well-formedness constraints on a program's execution.
  \item We propose a compositional proof system for this language
    capable of relating high-level application invariants to the structure
    of traces induced by the operational semantics, in which transactions
    are associated with specific weak-isolation levels.
  \item We define a \emph{maximum visibility principle} to reconcile
    store consistency guarantees with transaction isolation
    requirements, and use it to instantiate our operational model and
    the proof system for various kinds of (weakly consistent) data
    stores.
  \item Case studies modeled after real-world scenarios demonstrate
    the applicability and utility of our proof methodology.
\end{itemize}

>>>>

The paper makes the following contributions:
\begin{itemize}
  \item We develop a semantics for a core language equipped with
    weakly-isolated transactions, demonstrating that a general
    parametric isolation semantics can be expressed as
    well-formedness constraints on a program's execution.
  \item We propose a compositional proof system for this language
    capable of relating high-level application invariants to the structure
    of traces induced by the operational semantics, in which transactions
    are associated with specific weak isolation levels.
  \item We define a \emph{maximum visibility principle} to reconcile
    store consistency guarantees with transaction isolation
    requirements, and use it to instantiate our operational model and
    the proof system for various kinds of (weakly-consistent) data
    stores.
  \item Case studies modeled after real-world scenarios demonstrate
    the applicability and utility of our proof methodology.
\end{itemize}

### 2016-07-04 motivation.tex
typo fixes mostly

As an alternative, consider the execution of this transaction under
\emph{Read Committed} ({\sc rc}) isolation level, which is weaker than
{\sc ser}\footnote{{\sc rc} is in fact the default isolation level in
Postgres 9.5 and Oracle 11g databases.} An {\sc rc} transaction is
isolated from the writes of uncommitted transactions, and is therefore
free from \emph{dirty reads}~\cite{berenson} that read uncommitted
data. In the current example, {\sc rc} isolation admits the following
two executions (Fig.~\ref{fig:rc-ex}) on a strongly consistent ({\sc
sc}) store, such as a relational database:

\begin{figure}[!h]
\centering
\subcaptionbox {
  {\sc rc} Execution 1
  \label{fig:motiv-eg-1-a}
} [
  0.55\columnwidth
] {
  \includegraphics[scale=0.5]{Figures/motiv-eg-1-a}
}
%\hspace*{0.5in}
\subcaptionbox {
  {\sc rc} Execution 2
  \label{fig:motiv-eg-1-b}
}{
  \includegraphics[scale=0.5]{Figures/motiv-eg-1-b}
}
\caption{\small A possible execution of the program shown in Fig.~\ref{fig:motiv-eg-1} under
  a read committed isolation level.}
\label{fig:rc-ex}
\end{figure}
The figure depicts an execution as a series of read, write, and commit
operations.  The subscript of an operation indicates the transaction
that executes it. In the first execution, transaction \C{Wd1} reads
the current balance (\C{k}) and writes the new balance (\C{k-a1}), but
before it commits transaction \C{Wd2} executes and commits, writing
the new balance (\C{k-a2}). {\sc rc} isolation prevent \C{Wd2} from
witnessing the uncommitted writes of transaction \C{Wd1}.
Subsequently committing \C{Wd1} leads to the loss of \C{Wd2}'s updates
(the so called \emph{lost update anomaly}~\cite{berenson}), resulting
in an incorrect balance of \C{k-a1}. The second execution describes a
similar scenario with \C{Wd1} and \C{Wd2} exchanging their roles.
Clearly, {\sc rc} is an excessively weak isolation level for this
program because it loses the updates of one transaction, resulting in
the violation of the post-condition.  A stronger isolation level that
prevents lost updates is required. \iso{Snapshot Isolation} ({\sc
si})~\cite{berenson} fits this requirement; {\sc si} effectively
serializes transactions that update a shared data object by aborting
and re-executing a transaction if write-write conflicts are detected
during its commit.\footnote{{\sc si} however does not serialize
transactions in the absence of write-write conflicts.} Since {\sc si},
unlike {\sc ser}, does not rely on expensive lock-based concurrency
control, it is also more efficient, making it appropriate for both
\C{Wd} transactions.

@@

First, we note that the example in Fig.~\ref{fig:motiv-eg-1} is a
concurrent program, and hence is amenable to \emph{rely-guarantee}
style reasoning~\cite{rgjones}, a compositional proof technique that
allows us to reason about the behavior of individual threads by
abstracting away interferences induced by other threads (collectively
called \emph{the environment}) into a \emph{rely} relation. In an
ordinary concurrent program, every environment step is a valid
interference in the current thread. However, in the presence of
transactions executing under various levels of weak isolation,
determining what constitutes an interference is a non-trivial problem.
For example, a \iso{Serializable} transaction admits no interference,
whereas within a transaction executing under \iso{Read Uncommitted}
isolation all interferences are valid. Between these two extremes are
various levels of isolation that admit some interferences while
prohibiting others. For example, \iso{Read Committed} isolation admits
interference of a committed transaction, but not that of an
uncommitted transaction. \iso{Snapshot Isolation} admits interference
from committed non-conflicting transactions, and so on.

For the rely-guarantee approach to be useful, it has to thus constrain
allowed interference in accordance with the chosen level of isolation
associated with a transaction. The first observation we make is that
we can constrain interference by constraining the definition of the
\emph{happens-before} ($\hbZ$) relation that dictates visibility
actions between transactions, or their respective constituents.  To do
so, we axiomatize the $\hbZ$ relation to capture the interference
characteristics of different isolation levels as their specifications.
Two such specifications are shown\footnote{Note that specifications
presented here are solely for illustration. The actual specifications
described in \S\ref{sec:ansi-isolation} are more nuanced.} in
Fig.~\ref{fig:interference-ex}.  \iso{Read Committed} specification
($\psi_{RC}$) allows an operation $\eta_1$ of $T_1$ to happen before
an operation $\eta_2$ of $T_2$ only if every operation in $T_1$
(including its commit) happens before $\eta_2$ (we let $T_1 \hboar
\eta_2$ denote this). The specification does not require $T_1$ to
execute and commit before all operations of $T_2$, thus allowing it to
interfere in $T_2$. The \iso{Snapshot Isolation} specification
($\psi_{SI}$) requires two transactions that write to the same
variable to be related by happens-before.  This effectively prohibits
interference due to actions in $T_1$ being interleaved in $T_2$ (or,
vise versa).

Fig.~\ref{fig:motiv-eg-1-hb} is a visualization of $\hbZ$ 
from \C{Wd2} to \C{Wd1} allowed by $\psi_{RC}$ and $\psi_{SI}$. All
arrows are legal under $\psi_{RC}$ because in no case does an operation
$\eta_1$ from \C{Wd1} happen-before $\eta_2$ of \C{Wd2} without the commit
of \C{Wd1} also happening before $\eta_2$. $\psi_{SI}$ however disallows
all dashed arrows, because dashed arrows establish $\hbZ$ between actions in \C{Wd2}
and only a subset of the operations in \C{Wd1}. Ligher (blue) arrows with
hollow heads denote $\hbZ$ relationships that do not affect the value
of \C{C} in a way that causes the program to violate its
post-condition.  Darker (black) arrows with solid heads denote $\hbZ$
relationships that lead to the violation of the post-condition. The task of
the reasoning framework is to determine if all $\hbZ$ relationships
allowed by an isolation level lead to the satisfaction of the
post-condition.

@@

The specification of an isolation level encodes its interference
characteristics as constraints over the $\hbZ$ relation. However, for
this to be useful in reasoning about programs, the rely-guarantee
framework should be able to use the specification to determine if an
interference is valid or invalid, allowing the programmer to only
focus on the former. Our second observation is that this is possible
if the reasoning framework adequately tracks $\hbZ$ at each program
point, while preserving $\hbZ$ constraints as invariants between
program points. An interference that leads to the violation of $\hbZ$
constraints (i.e., an invalid interference) is thus automatically
prohibited. For instance, consider the program point after the write
to \C{C} in \C{Wd1}. The expected invariant ($\phi$) at that program
point is shown below ($\committed$ stands for ``committed''):

\begin{smathpar}
\begin{array}{l}
  \neg\committed(\C{Wd1}) \conj (\neg\committed(\C{Wd2}) \Rightarrow
  \C{C = k-a1}) 
                \\
      \C{Wd1} \wrstoar \C{C} \conj (\committed(\C{Wd2})
                \Rightarrow \C{C = k-a1-a2})
\end{array}
\end{smathpar}

>>>>

As an alternative, consider the execution of this transaction under
a \emph{Read Committed} ({\sc rc}) isolation level, which is weaker than
{\sc ser}.\footnote{{\sc rc} is in fact the default isolation level in
Postgres 9.5 and Oracle 11g databases.} An {\sc rc} transaction is
isolated from the writes of uncommitted transactions, and is therefore
free from \emph{dirty reads}~\cite{berenson} that read uncommitted
data. In the current example, {\sc rc} isolation admits the following
two executions (Fig.~\ref{fig:rc-ex}) on a strongly consistent ({\sc
sc}) store, such as a relational database:

\begin{figure}[!h]
\centering
\subcaptionbox {
  {\sc rc} Execution 1
  \label{fig:motiv-eg-1-a}
} [
  0.55\columnwidth
] {
  \includegraphics[scale=0.5]{Figures/motiv-eg-1-a}
}
%\hspace*{0.5in}
\subcaptionbox {
  {\sc rc} Execution 2
  \label{fig:motiv-eg-1-b}
}{
  \includegraphics[scale=0.5]{Figures/motiv-eg-1-b}
}
\caption{\small A possible execution of the program shown in Fig.~\ref{fig:motiv-eg-1} under
  a read committed isolation level.}
\label{fig:rc-ex}
\end{figure}
The figure depicts an execution as a series of read, write, and commit
operations.  The subscript of an operation indicates the transaction
that executes it. In the first execution, transaction \C{Wd1} reads
the current balance (\C{k}) and writes the new balance (\C{k-a1}), but
before it commits, transaction \C{Wd2} executes and commits, writing
the new balance (\C{k-a2}). {\sc rc} isolation prevent \C{Wd2} from
witnessing the uncommitted writes of transaction \C{Wd1}.
Subsequently committing \C{Wd1} leads to the loss of \C{Wd2}'s updates
(the so-called \emph{lost update anomaly}~\cite{berenson}), resulting
in an incorrect balance of \C{k-a1}. The second execution describes a
similar scenario with \C{Wd1} and \C{Wd2} exchanging their roles.
Clearly, {\sc rc} is an excessively weak isolation level for this
program because it loses the updates of one transaction, resulting in
the violation of the post-condition.  A stronger isolation level that
prevents lost updates is required. \iso{Snapshot Isolation} ({\sc
si})~\cite{berenson} fits this requirement; {\sc si} effectively
serializes transactions that update a shared data object by aborting
and re-executing a transaction if write-write conflicts are detected
during its commit.\footnote{{\sc si} however does not serialize
transactions in the absence of write-write conflicts.} Since {\sc si},
unlike {\sc ser}, does not rely on expensive lock-based concurrency
control, it is also more efficient, making it appropriate for both
\C{Wd} transactions.

@@

First, we note that the example in Fig.~\ref{fig:motiv-eg-1} is a
concurrent program, and hence is amenable to \emph{rely-guarantee}
style reasoning~\cite{rgjones}, a compositional proof technique that
allows us to reason about the behavior of individual threads by
abstracting away interferences induced by other threads (collectively
called \emph{the environment}) into a \emph{rely} relation. In an
ordinary concurrent program, every environment step is a valid
interference in the current thread. However, in the presence of
transactions executing under various levels of weak isolation,
determining what constitutes an interference is a non-trivial problem.
For example, a \iso{Serializable} transaction admits no interference,
whereas within a transaction executing under \iso{Read Uncommitted}
isolation, all interferences are valid. Between these two extremes are
various levels of isolation that admit some interferences while
prohibiting others. For example, \iso{Read Committed} isolation admits
interference of a committed transaction, but not those from
uncommitted transactions. \iso{Snapshot Isolation} admits interference
from committed non-conflicting transactions, and so on.

For the rely-guarantee approach to be useful, it has to thus constrain
allowed interference in accordance with the chosen level of isolation
associated with a transaction. The first observation we make is that
we can constrain interference by constraining the definition of the
\emph{happens-before} ($\hbZ$) relation that dictates visibility
actions between transactions, or their respective constituents.  To do
so, we axiomatize the $\hbZ$ relation to capture the interference
characteristics of different isolation levels as their specifications.
Two such specifications are shown\footnote{Note that specifications
presented here are solely for illustration. The actual specifications
described in \S\ref{sec:ansi-isolation} are more nuanced.} in
Fig.~\ref{fig:interference-ex}.  The \iso{Read Committed} specification
($\psi_{RC}$) allows an operation $\eta_1$ of $T_1$ to happen-before
an operation $\eta_2$ of $T_2$ only if every operation in $T_1$
(including its commit) happens-before $\eta_2$ (we let $T_1 \hboar
\eta_2$ denote this). The specification does not require $T_1$ to
execute and commit before all operations of $T_2$, thus allowing it to
interfere in $T_2$. The \iso{Snapshot Isolation} specification
($\psi_{SI}$) requires two transactions that write to the same
variable to be related by happens-before.  This effectively prohibits
interference due to actions in $T_1$ being interleaved in $T_2$ (or,
vise versa).

Fig.~\ref{fig:motiv-eg-1-hb} is a visualization of $\hbZ$ 
from \C{Wd2} to \C{Wd1} allowed by $\psi_{RC}$ and $\psi_{SI}$. All
arrows are legal under $\psi_{RC}$ because in no case does an operation
$\eta_1$ from \C{Wd1} happen-before $\eta_2$ of \C{Wd2} without the commit
of \C{Wd1} also happening before $\eta_2$. $\psi_{SI}$ however disallows
all dashed arrows, because dashed arrows establish $\hbZ$ between actions in \C{Wd2}
and only a subset of the operations in \C{Wd1}. Lighter (blue) arrows with
hollow heads denote $\hbZ$ relationships that do not affect the value
of \C{C} in a way that causes the program to violate its
post-condition.  Darker (black) arrows with solid heads denote $\hbZ$
relationships that lead to the violation of the post-condition. The task of
the reasoning framework is to determine if all $\hbZ$ relationships
allowed by an isolation level lead to the satisfaction of the
post-condition.

@@

The specification of an isolation level encodes its interference
characteristics as constraints over the $\hbZ$ relation. However, for
this to be useful in reasoning about programs, the rely-guarantee
framework should be able to use the specification to determine if an
interference is valid or not, allowing the programmer to only
focus on the former. Our second observation is that this is possible
if the reasoning framework adequately tracks $\hbZ$ at each program
point, while preserving $\hbZ$ constraints as invariants between
program points. An interference that leads to the violation of $\hbZ$
constraints (i.e., an invalid interference) is thus automatically
prohibited. For instance, consider the program point after the write
to \C{C} in \C{Wd1}. The expected invariant ($\phi$) at that program
point is shown below ($\committed$ stands for ``committed''):\vspace*{-8pt}

\begin{smathpar}
\begin{array}{c}
 \neg\committed(\C{Wd1}) \conj \C{Wd1} \wrstoar \C{C}  \conj  \\
    (\neg\committed(\C{Wd2}) \Rightarrow \C{C = k-a1}) \conj (\committed(\C{Wd2})
                \Rightarrow \C{C = k-a1-a2})
\end{array}
\end{smathpar}
\vspace*{-8pt}
